---
title: "DHIS2 Scheduled SMS"
output:
  html_document: 
    toc: true
    toc_float: false
    toc_depth: 4
    number_sections: true
params:
  minify_DT: true
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo=FALSE, message=TRUE, warning=FALSE, paged.print=FALSE, fig.width = 7)
options(encoding = "UTF-8")

Sys.setlocale(category = "LC_ALL", "Arabic")
options(dplyr.summarise.inform=F) 


##Important!! Enter credential information. Include last "/" of the baseurl
##ENTER BASE URL, USERNAME, PASSWORD
baseurl<-"https://hrhr.dhis2.org/dhis/"
username<-""
password<-""

#specify ANC stage IDS to extract events from (improves performance)
stage_ids<-c("edqlbukwRfQ","WZbXY0S00lP")


##Load required packages
required_packages<-c("openair","ggplot2","dplyr","stringr","lubridate","httr","assertthat","readr","stats","graphics","datasets","DT","methods","base","knitr")   

is_installed <- function(mypkg) is.element(mypkg, installed.packages()[,1])
load_or_install<-function(required_packages) {
  for(package_name in required_packages)  {
    if(!is_installed(package_name))     {
      install.packages(package_name,repos="http://lib.stat.cmu.edu/R/CRAN")  }
    library(package_name,character.only=TRUE,quietly=TRUE,verbose=FALSE)
  }
}
load_or_install(required_packages)

  
##download libraries and test login
require(readr)
require(assertthat)
require(httr)
loginDHIS2<-function(baseurl,username,password) {
  url<-paste0(baseurl,"api/me")
  r<-GET(url,authenticate(username,password))
  assert_that(r$status_code == 200L) }
if(loginDHIS2(baseurl,username,password)){
  print("successfully logged in")
}else{
  stop("could not log in! Please check url, username and password")
}

```

* Analysis run on `r Sys.Date()`

* Pulling Scheduled Events and Scheduled SMS for `r baseurl`

```{r schedule, echo=FALSE, warning=FALSE, include=FALSE}

##pull program IDs
require(httr)
program_api<-"api/programs.csv?fields=id"
url<-paste0(baseurl,program_api)
program<-as.data.frame(read_csv(content(GET(url), as = "text")))


#create url query
events_function<- function(baseurl, program, stage){
    
              paste0(baseurl,
                   "api/events.json?fields=program,orgUnit,event,",
                    "programStage,created,status,enrollment,trackedEntityInstance,dueDate",
                    "&startDate=2019-01-01",
                    "&includeDeleted=TRUE",
                    "&orgUnit=GD7TowwI46c&ouMode=DESCENDANTS",
                    "&paging=false",
                    "&program=", program,
                    "&programStage=",stage) 
  }




#now join with stages
program<-data.frame(stage_id=stage_ids,
                    program_id=rep(program$id, length(stage_ids))) %>% 
          mutate(url=events_function(baseurl, program_id, stage_id))
                    


#for each program ID, pull scheduled events and add to a dataframe
#first, create an empty dataframe
sched_events<-tibble()
#e<-c("programStage","orgUnit","program","event","status","created")
#for (k in e) sched_events[[k]]<-as.character()

#now loop over program +stages to get events
for (value in 1:length(program$url)){

url<-program$url[value]
  
event1<-jsonlite::fromJSON(content(GET(url),"text"))
event1<-as.data.frame(event1$events)

if (length(event1$event)>0){
  sched_events<-rbind(sched_events,event1)
}
next()
}


#debug checks
#head(sched_events)
#length(unique(sched_events$event))
#now pull all scheduled SMS messages, and remove those that are based on program enrollment
sms_api<-"api/sms/scheduled"
url<-paste0(baseurl,sms_api)
sched_sms<-as.data.frame(jsonlite::fromJSON(content(GET(url),"text"), flatten = TRUE))
sched_sms<- sched_sms[!is.na(sched_sms$programStageInstance.id),]
sched_sms<-(subset(sched_sms, select=c("name", "created","scheduledAt","id","lastUpdatedBy.name", "lastUpdatedBy.id","programStageInstance.created","programStageInstance.id")))
colnames(sched_sms)<-c("msg_name", "msg_generated","msg_scheduledFor","msg_id","creater_name","creater_userid", "event_createdat","event_id")

```


* Looping through scheduled events at SMS/sent endpoint

```{r sent-failed SMS, echo=FALSE}
#create empty dataframe to capture all sent msgs
sent_sms<-data.frame(event_id =character(),
                     status =factor(),
                     phone = integer(),
                     msg_id = character(), 
                     processed = character())

#loop through scheduled SMS by event_id at: SENT API endpoint
#skip over events that return an error response (deleted event after scheduled msg), or those that return a null array (no sent messages)

require(httr)
unique_events<-unique(sched_sms$event_id)
for (i in 1:length(unique_events)){
psi<-unique_events[i]
sms_test<-"api/sms/scheduled/sent?programStageInstance="
url<-paste0(baseurl,sms_test,psi)
resp<-as.data.frame(suppressMessages(jsonlite::fromJSON(content(GET(url),"text"), flatten = TRUE)))

if (ncol(resp)==4){
  next()
}
if (nrow(resp)==0) {
  next()
}

  event_id<-as.character(psi)
  status<-as.character(resp$messageStatus)
  phone<-as.character(resp$recipients.phoneNumbers)
  msg_id<-as.character(resp$id)
  msg_text<-as.character(resp$text)
  processed<-as.character(resp$processedDate)
  if(is.null(resp$recipients.phoneNumbers)){
    phone<-"null"}
  sched_response<-as.data.frame(cbind(event_id, status, phone, msg_id, msg_text, processed))
  sent_sms<-rbind(sent_sms,sched_response)
}
#convert to character
sent_sms$phone<-as.character(sent_sms$phone)


```


```{r get org units, echo=FALSE, warning=FALSE, message=FALSE}
library(jsonlite)
library(dplyr)
library(readr)

url<-paste0(baseurl,"api/29/organisationUnits.csv?paging=false&fields=name,id")

all_OUs<-readr::read_csv(content(GET(url)))
               

#join with sched_events
sched_events<- sched_events %>% 
  left_join(all_OUs, by = c("orgUnit" = "id"))

```




```{r anonymize phone numbers, echo = FALSE, eval = TRUE}

#join sent SMS with events to get the due date on the message and TEI ID
 sent_sms2 <- sent_sms %>%
  mutate(event=as.character(event_id)) %>%
  left_join(sched_events, by =  c("event" = "event"))


#copy to output table, retain original for CSV
sent_sms_table<-sent_sms2 %>%
 mutate("Tracker_Event"=paste0("<a href=\"", baseurl,
                          "/dhis-web-tracker-capture/index.html#/dashboard?",
                          "tei=",trackedEntityInstance,
                          "&program=WSGAb5XwJ3Y&ou=",orgUnit,
                          "\">",event_id,"</a>")) %>%
select(Tracker_Event,name,msg_text,Send_Status="status.x", phone, processed, Status_Event="status.y", dueDate, created)


distinct_receivers<-length(unique(sent_sms2$trackedEntityInstance))


#replace digits with xxxx, keep last 4
library(stringr)
str_sub(sent_sms_table$phone, start = 1, end = nchar(sent_sms_table$phone)-4) <- "XXXX"

```


```{r extract date from sent SMS}
date_extracted <- sent_sms_table %>%
  select(Tracker_Event,name,processed,dueDate, msg_text) %>%
  mutate("sched_event_date" = as.Date(dueDate)) %>%
  mutate("msg_text_date" = as.Date(str_extract(msg_text, 
                                     "[0-9]{4}-[0-9]{2}-[0-9]{2}"), 
                                        format="%Y-%m-%d")) %>%
  mutate(msgdate_before_schedEvent = ifelse((
                                        is.na(msg_text_date) | is.na(sched_event_date)), 0,
                                        (ifelse((msg_text_date > sched_event_date), 2, 
                                        (ifelse((msg_text_date == sched_event_date), 1, 3))))))

mismatches<-date_extracted %>% 
  filter(msgdate_before_schedEvent >1)

```

```{r summarize outputs}

sched_sms_table<-sched_sms %>%
  group_by(msg_name) %>% 
  summarize("ScheduledSMS"=n()) %>%
  arrange(desc(ScheduledSMS))


sched_sms_byOU<-sched_sms %>%
  left_join(sched_events, by=c("event_id" = "event")) %>%
  group_by(name, msg_name) %>% 
  summarize("ScheduledSMS"=n()) %>%
  arrange(desc(ScheduledSMS))

library(lubridate)
dupl_sms_by_OU <-sent_sms2 %>%
  mutate(date_attempt = lubridate::date(processed)) %>%
  group_by(name, event, date_attempt, msg_text) %>%
  filter(n()>1) %>%
  summarize(Duplicates=n())


sched_events_table<-sched_events %>%
  filter(status=="VISITED" | status=="SCHEDULE" | status=="COMPLETED") %>%
  group_by(name, orgUnit) %>%
  summarize("ScheduledEvents"=n()) %>%
  arrange(desc(ScheduledEvents))



makeDT<-function(x, caption, pagelength){
DT::datatable(x, 
              caption = caption,
              escape = FALSE,
              extensions = 'Buttons', 
              options = list(
                pageLength = pagelength,
                dom = 'Blfrtip', 
                buttons = list(I("colvis"), c('copy', 'csv', 'excel','print'))))
  
}

```

# Summary of Data from script
CSV files can be found at `r getwd()`

`r distinct_receivers` distinct patients received SMS. In sum, the following scheduled SMS were successfully sent and failed from the DHIS2 instance.

```{r summaries, echo=FALSE, include=TRUE}
table(sent_sms$status)

write.csv(sent_sms, file="dhis_sent_scheduled_sms.csv")
write.csv(sched_sms, file="dhis_all_scheduled_sms.csv")
write.csv(sched_events, file="dhis_all_scheduled_events.csv")

```


## Sent SMS Table
```{r summaries 1}

if (params$minify_DT==TRUE){
  sent_sms_table <- sent_sms_table %>%
    select(-dueDate, -Tracker_Event) %>% 
    mutate(msg_text=paste0("...",
                           str_sub(msg_text, start=10, end=80),
                           "..."))
}


makeDT(sent_sms_table, "All SMS sent (blocked #s)", 5)
```

## Summary by Phone Number Prefix

Check here in case some providers (denoted by phone prefix) are failing to send SMS.

```{r Prefix, include=FALSE, echo=FALSE, fig.height=3, message=FALSE, warning=FALSE}

library(dplyr)
sent_sms$prefix<-stringr::str_sub(sent_sms$phone,1,3)
prefix_summary<-sent_sms %>%
  group_by(prefix, status) %>%
  summarize("count"=n()) %>%
  arrange(desc(count))

makeDT(prefix_summary, "Summary by Prefix", 5)

```

## Check for Duplicate Messages

For troubleshooting a known bug. 
These messages were sent to the same org unit, same event, same message, and sent on same date. 

```{r Duplicates, include=FALSE}

makeDT(dupl_sms_by_OU, "Duplicates", 5)

```

## Check for Messages With Date Mismatch

The date in the text message should match the exact date of the event's dueDate.
In total, there were `r length(mismatches$name)` mismatches.

```{r Date mismatch, include=FALSE}

makeDT(date_extracted, "Extracted Message Date", 5)

```



## All Scheduled Events, by Org Unit

The following table shows all `r sum(sched_events_table$ScheduledEvents)` Scheduled and Visited *EVENTS*


```{r summaries 2, include=TRUE}

#Scheduled Events by OU
makeDT(sched_events_table, "All Scheduled and Visited Events, by OU", 5)

```


## All SMS Scheduled, by Type
The next tables show all `r sum(sched_sms_table$ScheduledSMS)` messages *SCHEDULED*

```{r summaries 3}

makeDT(sched_sms_table, "Scheduled templates", 5)
```


## All SMS Scheduled, by Type and By Org Unit
```{r summaries 4}

makeDT(sched_sms_byOU, "Scheduled templates by org unit", 5)

```


# Graphs and Figures

```{r graphs1, echo = FALSE, fig.height= 6, message=FALSE, warning=FALSE}
ggplot(sched_sms_table, aes(x= reorder(msg_name, ScheduledSMS), y=ScheduledSMS))+
  ggtitle("Scheduled SMS by type") +
  geom_bar(stat="identity", width=0.8, fill="steelblue")+
  coord_flip() +
  theme_minimal() +
    theme(
    legend.title=element_blank(),  
    axis.title.y=element_blank())
```

```{r change language, echo= FALSE, message=FALSE, results= 'hide'}
Sys.setlocale(category = "LC_ALL", "English")

```


```{r figures2, echo = FALSE, fig.height=4, fig.width=11, message=FALSE, warning=FALSE}
#calendar view of SMS messages sent
require(openair)
require(lubridate)
require(stringr)
require(dplyr)
require(ggplot2)

sent_sms$day<-lubridate::mday(sent_sms$processed)
sent_sms$month<-as.factor(lubridate::month(sent_sms$processed))
sent_sms$year<-as.factor(lubridate::year(sent_sms$processed))
sent_sms$date<-lubridate::date(sent_sms$processed)


library(openair)
redcols<-openair::openColours(c("pink","coral","firebrick"), 15)
greencols<-openair::openColours(c("lightgreen","green","forestgreen"), 15)
date_summary<-sent_sms %>%
  group_by(date, status) %>%
  summarize(n())
colnames(date_summary)<-c("date","status","number")

summ_sent<- date_summary %>% filter(status=="SENT")
summ_failed<-date_summary %>% filter(status=="FAILED")
if (length(summ_sent$number)>0){
summary(summ_sent)
openair::calendarPlot(summ_sent, pollutant="number", xlab="Dates SMS were sent", cols=greencols) 
}


if (length(summ_failed$number)>0){
  summary(summ_failed)
  openair::calendarPlot(summ_failed, pollutant = "number", xlab="Dates SMS failed", cols=redcols)
}

##sent/failed by date (ugly line plot)
require(ggplot2)
ggplot(date_summary, aes(x=date, y=number, group=status))+
  geom_line(aes(color=status)) +
  geom_point(aes(color=status)) +
  labs(title="SMS send attempts by date")
##sent/failed by month (cleaner barplot)
ggplot(sent_sms, aes(month,fill=status)) +
         geom_bar(stat="count") +
         facet_wrap(~year) +
        labs(title="SMS send attempts by year and month")

#plot by messages schedled (includes forthcoming)
sched_sms$week<-lubridate::isoweek(sched_sms$msg_scheduledFor)
sched_sms$year<-lubridate::year(sched_sms$msg_scheduledFor)
sched_sms$yearweek<-paste0(sched_sms$year,"-",sched_sms$week)
#Set line for today's week
wtoday<-lubridate::week(Sys.Date())
year<-lubridate::year(Sys.Date())
today<-as.data.frame(cbind(wtoday,year))
ggplot(sched_sms, aes(week)) +
  geom_bar(stat="count") +
  facet_wrap(~year) +
  geom_vline(data=today, aes(xintercept=wtoday)) +
  geom_text(data=today, mapping=aes(x=wtoday, y=0, label="Current Week"), size=4, angle=90, vjust=-.4, hjust=-.4) +
  labs(y = "Total SMS scheduled", title ="Scheduled SMS by Year and Week")


```

End analysis